Lyman-$\alpha$ photons through rotating outflows

### 2018-03-31 paper/paper.tex
results

Recent improvements in instrumentation have revolutionized the kind of
kinematic studies that can be performed on galaxies with small angular
sizes on the sky.
For instance, spatially resolved spectra on dwarf galaxies show signs
of coherent rotation \citep{2009A&A...493..871S}.
Some Compact Dwarf Galaxies actually show gas kinematics that range
from pure rotation to high velocity dispersion without a clear
rotation pattern.
\citep{2015A&A...577A..21C,2017A&A...600A.125C}
These studies are now being extended to distant galaxies.

Some of those galaxies are expected to have high neutral gas contents
and thus show \lya\ emission with rotation imprints
\citep{2005A&A...433L...1B,2008ApJ...672..888T,2013MNRAS.434.2491G}
The computational modeling of systems showing rotation has only been
studied until recently by \cite{Garavito14}.
In that work the authors studied the influence of pure solid body
rotation on the \lya line's morpholgy.
They found that rotation indeed introduces changes, the most
noticeable change is the dependence of the spectra with the viewing
angle with respect to the rotation axis.

\cite{Garavito14} also presented a simple semi-analytical
approximation that accounted for the main features of the \lya spectra
from a rotation sphere.
Recently, this semi-analytic solution was used to perform a Markov
Chain Monte Carlo exploration to fit the observed spectra Compact
Dwarf Galaxy \citep{tololo}, that presented atypical features that
could be explained by pure rotation.

@@

In Table \ref{tab:values} we list the combination of \tauh,
\vrot and \vrot values used in this paper.
The range of values have some overlap with the expectations from a
galaxy with a total neutral hydrogen mass of $10^8$-$10^9$
$M_{\odot}$. We run a total of $27$ different models.

@@

\subsubsection{Match to \lya observations by MUSE}

In order to give the reader a feeling of the ranges of observational
data that can be explained by the models presented here we perform a
comparison against observational data from MUSE.

@@

The main effects of rotation on the \lya\ line morphology are:

\begin{itemize}
  \item Broadening the line.
	\item Increasing the intensity at the line's center.
	\item Inducing a dependency on the viewing angle,
  The closer the observer is to the pole of the galaxy (defined by the rotation axis),
  the smaller are the effects from rotation.
\end{itemize}

We found that in the case of solid body rotation these effects can be quantitatively
explained by a Doppler boost, where the Doppler factor can be computed as the
 producet of quantities at the surface of last scattering, namely $\vec{v}_{\rm rot}\cdot \hat{k}$,
 where $\vec{v}_{\rm rot}$ is the velocity due to rotation and $\hat{k}$ is the direction
of the photon's propagation.
These conclusions, specially the confirmation of the Doppler boost as
a good proxy for solid body rotation, strenghten the evidence reported
by \cite{Garavito14} when they modeled the effect of pure rotation on
\lya\ spectra.

As an application to observational data we find that recent results
that take the spectra of two different sides of a galaxy can detect 
the effect of approaching/receeding gas motions. 
However, the distances between the peaks of these two spectra 
correspond to $\approx v_{\rm rot} \cos\theta$ due to the weights of the
emitting regions.

@@

\appendix

>>>>

Recent improvements in instrumentation have revolutionized the kind of
kinematic studies that can be performed on galaxies with small angular
sizes on the sky.
For instance, spatially resolved spectra on dwarf galaxies show signs
of coherent rotation \citep{2009A&A...493..871S}.
Some Compact Dwarf Galaxies actually show gas kinematics that range
from pure rotation to high velocity dispersion without a clear
rotation pattern.
\citep{2015A&A...577A..21C,2017A&A...600A.125C}
Some of those star forming galaxies are expected to have high neutral
gas contents and low dust contents and thus show \lya\ emission with
rotation imprints
\citep{2005A&A...433L...1B,2008ApJ...672..888T,2013MNRAS.434.2491G}

The computational \lya modeling of systems showing rotation has only
been studied until recently by \cite{Garavito14}.
In that work the authors studied the influence of pure solid body
rotation on the \lya line's morpholgy.
They found that rotation indeed introduces changes, the most
noticeable change is the dependence of the spectra with the viewing
angle with respect to the rotation axis.

\cite{Garavito14} also presented a simple semi-analytical
approximation that accounted for the main features of the \lya spectra
from a rotation sphere.
Recently, this semi-analytic solution was used to perform a Markov
Chain Monte Carlo exploration to fit the observed spectra Compact
Dwarf Galaxy \citep{tololo} with atypical features that could be
explained by pure rotation.

@@

As input parameters we use \tauh$=\{10^5, 10^6, 10^7\}$,
\vout$=\{5,25,50\}\kms$ and \vrot$=\{0,50,100\}\kms$, for a total of
$27$ models with all the possible parameter combinations.
The range of values have some overlap with the expectations from a
galaxy with a total neutral hydrogen mass of $10^8$-$10^9$
$M_{\odot}$.

@@

\subsubsection{Relationship to MUSE observations}

In order to give the reader a feeling of the ranges of observational
data that can be explained by the models presented here we perform a
comparison against observational data from MUSE-Wide survey (MWS)
\citep{2017A&A...606A..12H}.

MWS provides public access to 831 spectra out of which 237 contain a
detected \lya line as dominante. 
The public dataset reports as the redshift of the galaxy the position
of the dominant peak.
We use this redshift to convert the spectra to rest-frame and compute
the standard deviation and skewness statitstics. 
Given the uncertainty on the galaxy redshift we do not compute the
valley/peak ratio. 
We do not try to perform a full fit to all the spectra but only to
explore whether the region of parameter space explored in our RT
simulations shows any crossing with the results from the MWS.

@@

We find that the main effects of rotation on the \lya\ line morphology are:

\begin{itemize}
  \item Inducing a dependency on the viewing angle.
  \item Broadening the line.
  \item Increasing the intensity at the line's center.
\end{itemize}

We find that these effects can be quantitatively explained by a
Doppler shift computed as product of quantities at the surface of last
scattering, namely  $\vec{v}_{\rm rot}\cdot \hat{k}$,  where
$\vec{v}_{\rm rot}$ is the velocity due to rotation and $\hat{k}$ is
the direction of the photon's propagation.  
These conclusions, specially the confirmation of the Doppler Boost as
a good approximation to model solid body rotation, strenghten the
semi-analytic model reported by \cite{Garavito14} when they modeled
the effect of pure rotation on \lya\ spectra.

As an application to observational data our results suggest that
rotation should also be considered as an explanation of the
non-vanishing \lya intensity at the line's center.
We also show how the \lya spectra from two different sides of a galaxy
can detect approaching/receeding gas motions We show that the
distances between the peaks of these two spectra is smaller than the
expected value of $2\times\vrot$ due to the differente weights at the
surface of the emitting regions.  
Finally we also use data from MUSE to demonstrate to what extent the
outputs of the RT results presented here can be expected to match
hundreds of observed \lya spectra.

Our work confirms that rotation produces features in the \lya line
strong enough that can be expected to be detected in observations and
that should be taken into account into \lya modelling; the Doppler
Boost offers an easy-to-implement approximation to explore such
influence into already existing RT simulations.

@@

\newpage
\appendix
